---
title: "P8157 HW3"
author: "Anna Ma"
date: "2022-10-30"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(data.table)
library(geepack)
library(gee)
library(lme4)
library(nlme)
library(doBy)
```

## Question 1 \ 

Randomized, double-blind, parallel-group, multicenter study comparing two oral treatments (denoted A and B) for toe-nail infection, patients were
evaluated for the degree of onycholysis (the degree of separation of the nail plate from the nail-bed) at baseline (week 0) and at weeks 4, 8, 12, 24, 36, and 48 thereafter. The onycholysis outcome variable is binary (none or mild versus moderate or severe). The binary outcome was evaluated on 294 patients comprising a total of 1908 measurements. 

The main objective of the analyses is to compare the effects of oral treatments A and B on changes in the probability of the binary onycholysis outcome over the duration of the study. 

* The *binary onycholysis outcome variable Y* is coded 0 = none or mild, 1 = moderate or severe.

* The categorical variable Treatment is coded 1=oral treatment A, 0=oral treatment B.

* The variable Month denotes the exact timing of measurements in months.

* The variable Visit denotes the visit number (visit numbers 1-7 correspond to scheduled visits at 0, 4, 8, 12, 24, 36, and 48 weeks).

```{r Q1 data import, message = FALSE}
toenail_df = read_delim(file = "toenail.txt", delim = " ", col_names = c("id", "response", "treatment", "month", "visit"), skip = 1) %>% mutate(treatment = factor(treatment))
```

##### 1. Consider a random effects model with a random intercept for the log odds of moderate or severe onycholysis. Assuming linear trends and month as the time variable.


* Setup: Since the response variable is binary such that $Y_{ij} = [0,1]$, we can assume that $y_{ij} \sim binomial(n, p_{ij})$, and that the mean response model is $log(\frac{\mu_{ij}}{1-\mu{ij}})$

* Random Effects Model with random intercept: 
    * $y_{ij} = X_{ij}\beta+b_{i}+\epsilon_{ij}$
    * $b_i \sim N(0, \sigma_{b}^2)$
    * $\epsilon_i \sim N(0, \sigma^2)$
    * $\epsilon_{ij}$ and $b_i$ are independent 

* Mean:
    * Population average: $logit(E[Y_{ij}]) = X_{ij}\beta$
    * Subject specific mean: $logit(E[Y_{ij}|b_i]) = X_{ij}\beta + b_i$

* Variance and Covariance: 
    * $Var[Y_{ij}] = \phi v(\mu_{ij}) = \mu_{ij}(1-\mu_{ij})$ 
    * Total Variance $Var[Y_{ij}] = \sigma_{b}^2 +\sigma^2$
    * Within Subject Variance $Var[Y_{ij}|b_i] = \sigma^2$
    * Covariance: $Cov(Y_{ij}, Y_{ik}) = \sigma_{b}^2$

The model can be written as:
$$
\begin{split}
logit(E[Y_{ij}|b_{i}])
&= (\beta_{0} + b_{i}) + \beta_{1}Treatment_{i} + \beta_{2}Month_{ij} + + \beta_{3}Treatment_{i} * Month_{ij} \\
\end{split}
$$

```{r}
rme_q1 = glmer(response ~ treatment*month + (1|id), family = "binomial", data = toenail_df)
summary(rme_q1)
```


##### 2. Provide Interpretations for the fixed effects coefficients in your model. Interpret the random effect parameter.\newline

**For fixed effects:**

```{r}
coef_q1 = summary(rme_q1)
coef_q1$coefficients  %>% knitr::kable(digits = 3)
```

* $\beta_{0}$: the log odds of the onycholysis outcome for an average/typical individual in treatment B group at baseline is `r round(coef_q1$coefficients[1], 3)`. 

* $\beta_1:$ the log odds ratio of the onycholysis outcome for patients in treatment A versus patients in treatment group B at baseline is `r round(coef_q1$coefficients[2], 3)`.

* $\beta_2:$ the log odds ratio of the onycholysis outcome for every unit increase in month for a patient in treatment group B is `r round(coef_q1$coefficients[3], 3)`

* $\beta_3:$ the difference in the log odds ratio of the onycholysis outcome for a individual in treatment A group versus a individual treatment B group with one unit increase of month is `r round(coef_q1$coefficients[4], 3)`.

* $\beta_{2} + \beta_{3}$: the log odds ratio of the onycholysis outcome for every unit increase in month for a patient in treatment group A is `r round(coef_q1$coefficients[3] + coef_q1$coefficients[4], 3)`

**For random effect:**

```{r}
coef_q1$varcor %>% knitr::kable(digits = 3)
```

* Random effect parameter: The variance of the random effects $\sigma_b^2$ is 20.76, this suggest that there is variation in the random effects between individuals. 

* 95% Confidence Interval: $\hat{\beta}_{0}\pm 1.96*\sigma_b = -2.51 \pm 1.96*4.56 = [-11.45, 6.43]$. Therefore, 95% of the subjects have the intercept in the range [-11.45, 6.43].


##### 3. From the results of your analysis what conclusions do you draw about the effect of treatment on changes in the severity of onycholysis over time? Provide results that support your conclusions.\newline    


From the model,we see that the odds ratio of the outcome for patients in treatment A versus patients in treatment group B at baseline is `r round(exp(coef_q1$coefficients[2]), 3)`. And since the estimated coefficient for this interaction is -0.137, which is negative, the difference between the log odds ratio of the onycholysis outcome for a individual in treatment A group versus a individual treatment B grows in the negative direction over time, meaning that the effect of treatment increases over time. \  

However, from the model,we see that with a p-value of 0.65728, the treatment beta value is not significant. So, although the p-value for the interaction term is 0.048 and therefore significant for the response outcome, it's effect is likely to be driven by month instead of the treatment itself. 


##### 4. How are the interpretations different from the model in HW2.

* HW 2 uses a GEE model, which interprets the parameters as population average.

* This HW uses a mixed effect model, random effect with random intercept, which interprets the parameters on an individual-specific level. 


## Question 2

The Skin Cancer Prevention Study was a randomized, double-blind, placebo-controlled clinical trial of beta carotene to prevent non-melanoma skin cancer in high-risk subjects.A total of 1805 subjects were randomized to either placebo or 50 mg of beta carotene per day for 5 years. 

The main objective of the analyses is to compare the effects of beta carotene on skin
cancer rates. 

* The outcome variable Y is a count of the of the number of new skin cancers per year. 
* The categorical variable Treatment is coded 1=beta carotene, 0=placebo. 
* The variable Year denotes the year of follow-up. 
* The categorical variable Gender is coded 1 male, 0 female. 
* The categorical variable Skin denotes skin type and is coded 1 = burns, 0 otherwise. 
* The variable Exposure is a count of the number of previous skin cancers.
* The variable Age is the age (in years) of each subject at randomization.

```{r Q2 data import,message = FALSE}
skin_df = read.table(file = "skin.txt", header = FALSE,col.names = c("id", "center", "age", "skin", "gender", "exposure", "y", "treatment", "year"), skip = 3) %>% mutate(treatment = factor(treatment))
```

##### 1. Set up a suitable random effects (random intercept) model for rate of skin cancers with Treatment and Year as covariates.

* Setup: Since the response variable is a count of the number of new skin cancers per year, we can assume that $y_{ij} \sim poisson(\lambda_{ij})$, and that the mean response model is $log(\mu_{ij}) =\eta_{ij}$

* Random Effects Model with random intercept: 
    * $y_{ij} = X_{ij}\beta+b_{i}+\epsilon_{ij}$
    * $b_i \sim N(0, \sigma_{b}^2)$
    * $\epsilon_i \sim N(0, \sigma^2)$
    * $\epsilon_{ij}$ and $b_i$ are independent 

* Mean:
    * Population average: $log(E[Y_{ij}]) = X_{ij}\beta$
    * Subject specific mean: $log(E[Y_{ij}|b_i]) = X_{ij}\beta + b_i$

* Variance and Covariance: 
    * $V(\mu_{ij})= \mu_{ij}$
    * Total Variance $Var[Y_{ij}] = \sigma_{b}^2 +\sigma^2$
    * Within Subject Variance $Var[Y_{ij}|b_i] = \sigma^2$
    * Covariance: $Cov(Y_{ij}, Y_{ik}) = \sigma_{b}^2$

The model can be written as:
$$
\begin{split}
log(E[Y_{ij}|b_{i}])
&= (\beta_{0} + b_{i}) + \beta_{1}Treatment_{i} + \beta_{2}Year_{ij} + + \beta_{3}Treatment_{i} * Year_{ij} \\
\end{split}
$$

```{r}
rme_q2 = glmer(y ~ treatment*year + (1|id), family = "poisson", data = skin_df)
summary(rme_q2)
```

##### 2. Provide Interpretations for the fixed effects coefficients in your model. Interpret the random effect parameter.\newline 

**For fixed effects:**

```{r}
coef_q2 = summary(rme_q2)
coef_q2$coefficients  %>% knitr::kable(digits = 3)
```

* $\beta_{0}$: For an average/typical patients in the placebo group, the log rate of the number of new skin cancers at baseline is `r round(coef_q2$coefficients[1], 3)`.

* $\beta_1:$ The log rate of the number of new skin cancers
for a patient in the treatment group versus a patient in the placebo group is `r round(coef_q2$coefficients[2], 3)`.

* $\beta_2:$ The log rate of the number of new skin cancers for one unit increase in year for a patient in the placebo group is `r round(coef_q2$coefficients[3], 3)`.

* $\beta_3:$ The difference in the log rate of the number of new skin cancers for one unit increase in year between a patient in treatment group versus a patient in the placebo group is `r round(coef_q2$coefficients[4], 3)`

* $\beta_2 +\beta_3$: The log rate of the number of new skin cancers for one unit increase in year in a patient in the treatment group is `r round(coef_q2$coefficients[3]+coef_q2$coefficients[4], 3)`

**Random effect:**

```{r}
coef_q2$varcor %>% knitr::kable(digits = 3)
```

* Random effect parameter: The variance of the random effects $\sigma_b^2$ is 2.189, this suggest that there is not a lot variation in the random effects between individuals. 

* 95% Confidence Interval: $\hat{\beta}_{0}\pm 1.96*\sigma_b = -2.415 \pm 1.96*1.48 = [-5.32, 0.486]$. Therefore, 95% of the subjects have the intercept in the range [-5.32, 0.486].


##### 3. From the results of your analysis what conclusions do you draw about the effect of beta carotene on the rate of skin cancers? Provide results that support your conclusions.\newline 


From the model,we can see that the p-value for the treatment covariate is 0.564, and that the p-value for the interaction term of treatment and year is 0.331. Both of the p-value is greater than the significance level of 0.05, therefore we fail to reject the null and conclude that the effect of both covariates are not significant for the response outcome.Thus, there's not enough evidence to show that beta carotene have a significant effect on the rate of skin cancers.  

##### 4. Repeat the above analysis adjusting for skin type, age, and the count of the number of previous skin cancers. What conclusions do you draw about the effect of beta carotene on the adjusted rate of skin cancers?

```{r, message = FALSE}
rme_q2_2 = glmer(y ~ treatment*year + skin + age + exposure + (1 | id), family = "poisson", data = skin_df)
coef_q22 = summary(rme_q2_2)
coef_q22$coefficients  %>% knitr::kable(digits = 3)
```

From the model, we can see that the p-value for the treatment covariate is 0.831, and that the p-value for the interaction term of treatment and year is 0.310. Both of the p-value is greater than the significance level of 0.05, therefore we fail to reject the null and conclude that the effect of both covariates are not significant for the response outcome.Thus, like the previous model, there's not enough evidence to show that beta carotene have a significant effect on the rate of skin cancers. On the other hand, covariates skin, age, and exposure have significant effect on the outcome response. 


##### 5. How are the interpretations different from the model in HW2.
* HW 2 uses a GEE model, which interprets the parameters as population average.

* This HW uses a mixed effect model, random effect with random intercept, which interprets the parameters on an individual-specific level. 
